FlowFPX: Nimble Tools for Debugging Floating-Point Exceptions

### 2023-08-01 paper.tex
quick notes + links
sec2 note: gpu-fpx has a great fp intro
Bunch of SVGs for the talk—update rough TOC for writing
Lots of work on §3

% If using Emacs, put cursor on last closing paren and call "eval-last-sexp" (usually `C-x C-e') to regenerate
%\iffalse
%(save-excursion
%  (let* ((toc (funcall imenu-create-index-function))
%         (headers (mapcar (pcase-lambda (`(,str . ,loc)) str)
%                          toc)))
%    (beginning-of-buffer)
%    (search-forward (concat "%% " "--- autogenerated TOC ---"))
%    (next-line)
%    (beginning-of-line)
%    (let ((start (point)))
%      (search-forward (concat "%% " "--- end autogenerated TOC ---"))
%      (beginning-of-line)
%      (delete-region start (point))
%      (goto-char start))
%    (cl-loop for header in headers
%             do (insert (format "%% %s\n" header)))
%    (next-line)))
%\fi

%% --- autogenerated TOC ---
% juliacon
% document
%   Introduction
%   Floating Point Exception Primer
%     Where do NaNs come from?
%     Infinity
%     To Kill a Floating Point
%   \FlowFPX{
%     \FT{
%     Catching a NaN Killer
%     NaN propagation monitoring
%     NaN injection
%   Case Studies
%     Surprise NaNs: ShallowWaters
%       CSTGs help us get a bigger picture
%     NaN fuzzing to find a kill: OrdinaryDiffEq
%     Finch
%     A Bug-Hunting Safari
%     \GPUFPX{
%   Related Work
%   Discussion
%     Tracking more than NaNs and Infs
%     Enhanced Fuzzing
%     Static FPX, Gradual Types
%   Acknowledgments
% document
%% --- end autogenerated TOC ---

@@

@@

FILL intro: float tracker, viz tools, gpu

@@

\FT{} works by leveraging Julia's type-based method dispatch system to capture everything from calls to the standard library down to basic arithmetic operations.
We do this by creating custom types \texttt{TrackedFloat16}, \texttt{TrackedFloat32}, and \texttt{TrackedFloat64} that wrap their corresponding \texttt{Float16}, \texttt{Float32} etc. counterparts.
We then defined methods for every function in the \texttt{Base} module, including new implementations for operators like \texttt{+} and \texttt{*}.
These new methods had to be defined for every combination of tracked/untracked arguments;
fortunately Julia's metaprogramming capabilities make this relatively simple.

The surrogate methods in place, every operation on a TrackedFloat can be intercepted.
We have three primary goals when intercepting a float:

\begin{enumerate}
  \item Catch instances of \NaN{} kills
  \item Monitor the propagation of \NaN{}s through the program
  \item Optionally \emph{inject} \NaN{}s to fuzz the program under scrutiny against \NaN{}-kills
\end{enumerate}

% Good things to put in this section:
% - start with the max example, perhaps?
% - trace examples
% - CSTGs
%
% Don't forget to talk about trace recording

\subsubsection{Catching a \NaN{} Killer}

With a \fp{} value wrapped in our \texttt{TrackedFloat} type, it is easy to monitor when a \NaN{} gets killed.
Julia dispatches all arithmetic operations on that value to flow through FloatTracker, where we can inspect the arguments.
Kills are relatively simple: if one or more \NaN{}s flow into a function but don't flow out (i.e. the answer isn't a \NaN{}) then that's a kill.
For example:

\begin{lstlisting}[language = Julia]
  NaN + 42     -> NaN    # not a kill
  Inf - Inf    -> NaN    # gen, but not a kill
  min(NaN, 42) -> NaN    # not a kill
  NaN < 42     -> false  # NaN kill!
\end{lstlisting}

There are a few exceptions to the rule of ``\NaN{} in and no \NaN{} out $\rightarrow$ kill''---some instances that we report as being a \NaN{} kill might be instead a \emph{\NaN{}-repair}, i.e. where the user explicitly checks for a \NaN{} and gets some non-\NaN{} value as a result.
Once clear instance of this is the \texttt{isnan} function.
Calling \texttt{isnan} and getting a boolean is expected behavior, and is usually used in error handling or in \emph{explicit} handling of unexpected values.
FloatTracker still reports these cases in the kill logs so as to be complete---the user must not blindly assume that every instance of a \NaN{} kill is bad and might be a case of \NaN{} repair.

% FIXME: we don't (yet) talk much about NaN repair in the rest of the paper.

% TODO?

\subsubsection{\NaN{} propagation monitoring}

Remember in our first example with the \texttt{bad\_max} and \texttt{good\_max} functions how we didn't have a way to tell \emph{where} the \NaN{}s originally came from?
That can be a tedious task.
However, FloatTracker makes this a lot easier by watching the life cycle of a \NaN{}.

% TODO

% FIXME: don't forget to talk about CSTGs

\subsubsection{\NaN{} injection}

% TODO

@@

% Local Variables:
% jinx-local-words: "CFL CSTGs Courant Friedrichs JuliaHub Lewy OrdinaryDiffEq RxInfer ShallowWaters TrackedFloat isnan"
% End:

>>>>

%% extra links:
%% - GPU Julia: https://cuda.juliagpu.org/stable/tutorials/introduction/
%% - GPU x2: https://nextjournal.com/sdanisch/julia-gpu-programming
%% - NaN vs missing https://discourse.julialang.org/t/is-there-any-reason-to-use-nan-instead-of-missing/84396

% If using Emacs, put cursor on last closing paren and call "eval-last-sexp" (usually `C-x C-e') to regenerate
%% (save-excursion
%%  (let* ((toc (funcall imenu-create-index-function))
%%         (headers (mapcar (pcase-lambda (`(,str . ,loc)) str)
%%                          toc)))
%%    (beginning-of-buffer)
%%    (search-forward (concat "%% " "--- autogenerated TOC ---"))
%%    (next-line)
%%    (beginning-of-line)
%%    (let ((start (point)))
%%      (search-forward (concat "%% " "--- end autogenerated TOC ---"))
%%      (beginning-of-line)
%%      (delete-region start (point))
%%      (goto-char start))
%%    (cl-loop for header in headers
%%             do (insert (format "%% %s\n" header)))
%%    (next-line)))

%% --- autogenerated TOC ---
% --
% --
% <<header.tex
% Introduction
% Floating Point Exception Primer
% . Where do \NaN{}s come from?
% . . TODO good quote, try to work this in
% . Infinity
% . To Kill a Floating Point
% \FlowFPX{}
% . \FT{}
% . . Catching a \NaN{} Killer
% . . \NaN{} propagation monitoring
% . . \NaN{} injection
% . Visualization
% . \GPUFPX{}
% Case Studies
% . Surprise \NaN{}s: ShallowWaters
% . . CSTGs help us get a bigger picture
% . \NaN{} fuzzing to find a kill: OrdinaryDiffEq
% . Finch
% . A Bug-Hunting Safari
% . \GPUFPX{}: Oceananigans
% Related Work
% Discussion
% . Tracking more than \NaN{}s and \Inf{}s
% . Enhanced Fuzzing
% . Static FPX, Gradual Types
% Acknowledgments
% <<bib.tex.tex
% --
%% --- end autogenerated TOC ---

@@

%% TODO use the GPU-FPX paper as a guide for the background. Give a little more detail than that, but not much!

@@

\FlowFPX{} is a growing set of tools to aid in tracking down \fp{} exceptions.
The two primary tools we cover in this paper are \FT{} and CSTG:

\begin{itemize}
\item \FT{}'s role is twofold: to trace the lifetime of exceptional \fp{} values and to enable inside-out fuzzing.
\item CSTG's role is to visualize the tracing output from \FT{} in compact form.
\end{itemize}

\GPUFPX\cite{llsflg-hpdc-2023} is another component of \FlowFPX{} that can track \fp{} exceptions inside a GPU kernel.
Please see the cited paper for more details on \GPUFPX.
Work to use \GPUFPX in tandem with \FT{} is ongoing.

%% TODO julia proposed consistent NaN handling,
%%  do we propagate the exact NaN
%%  https://github.com/JuliaLang/julia/issues/48523
%%
%% Ashton: @Ben we simply propagate whatever we got from the operation. If
%% Julia's behavior around an operator change, that should be preserved even
%% when using FloatTracker

@@

\FT{} has two roles:

\begin{enumerate}
  \item Track exceptional values
  \item Fuzz \fp{} code from the inside-out
\end{enumerate}

Building the mechanism to track exceptional values also gave us a way to inject exceptional values wherever \fp{} operations occur.
We will examine each of these roles, how they are used, and how \FT{} realizes them.

%% What it does, then how it works

\subsubsection{Tracking Excptional Values}

\FT{}'s first purpose is to monitor \fp{} operations for exceptional values.
\FT{} interposes on every arithmetic operation and standard library call that involves a \fp{} and logs a stack trace to that point whenever an exception is generated, propagated, or killed.
A programmer can then use these stack traces to pinpoint where an exceptional value like \NaN{} came into being or where a guard against a \NaN{} kill is needed.
\FT{} can also log the arguments to the suspect call, so you can get \emph{normal, concrete} values to help you further discover what is going wrong.

All the logs that \FT{} captures are straight from Julia's \texttt{stacktrace()} function.
\FT{} writes these logs to file; one file for each kind of event.
This way, if a user of \FT{} is attempting to determine where a \NaN{} or an \Inf{} came from, they can look in the file for all the gen events.
Likewise, if the user wants to find where \NaN{}s get killed, then they can look in the kills file.

While other tools like profilers might accomplish this tracking by writing extensions to the language's compiler or runtime, we were able to implement \FT{} entirely in user land.
We created custom types \texttt{TrackedFloat16}, \texttt{TrackedFloat32}, and \texttt{TrackedFloat64} that wrap their corresponding \texttt{Float16}, \texttt{Float32}, and \texttt{Float64} counterparts.
We then defined methods for every function in the \texttt{Base} module, including new implementations for operators like \texttt{+} and \texttt{*}.
This lets us intercept all \fp{} operations working on our \texttt{TrackedFloat} types.

\begin{lstlisting}[language = Julia]
function Base.+(x::TrackedFloat64, y::TrackedFloat64)
  result = x.val + y.val
  check_error(+, result, x.val, y.val)
  TrackedFloat64(result)
end
\end{lstlisting}

These new methods had to be defined for every combination of tracked/untracked arguments;
fortunately Julia's metaprogramming abilities made it simple for us to scale without the code becoming unwieldy.
We made a macro that first loops through all the \texttt{TrackedFloat} types that we wished to implement, and then looped through all of the operators that we wished to support.
This approached saved us over 10x the lines a non-macro approach would have required.\footnote{We implemented overrides for 645 function/type combinations. At roughly 5 lines per variant, this is 3225 lines of code. Our implementation at time of writing weighs in at 218 lines of code, which is \emph{including} whitespace and comments.}
This approach was inspired by \texttt{Sherlogs.jl}.\cite{kMilanklSherlogsJl2021}

With the overrides in place, we can examine the arguments to a \fp{} operation to determine if an interesting event occurred.
Here is how we classify events around a \NaN{}:

\begin{tabular}{ll}
  Arguments & Result & Event Type \\
  \hrule
  No \NaN{} & \NaN{} & Gen \\
  \NaN{} & \NaN{} & Prop \\
  \NaN{} & No \NaN{} & Kill \\
\end{tabular}

The same classification rules apply to \Inf{} and would apply to any other value of interest.

\subsubsection{Fuzzing}

The second role of \FT{} is to fuzz code from the inside out.
While it is wholly possible to fuzz a program by giving it random inputs, this might not help you find \emph{all} the places where an exceptional value could crop up during a calculation.\cite{ddghlllprr-correctness-2022}

The exception tracking infrastructure that we have gives us a simple way to inject exceptional values wherever we want in our code.
We have a global configuration system that allows us to turn on fuzzing.
Then, each time we intercept a \fp{} operation, we check to see if we should inject a \NaN{}.\footnote{At time of writing, \FT{} only injects \NaN{}, but it is trivial to extend the set of values to inject.}
Our full decision process is as follows:

\begin{enumerate}
\end{enumerate}

@@

% Local Variables:
% jinx-local-words: "CFL CSTG's CSTGs Courant Friedrichs JuliaHub Lewy OrdinaryDiffEq RxInfer ShallowWaters TrackedFloat isnan"
% End:
